Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-04-30 _tex/paper2.tex
Built site for gh-pages

Nevertheless, despite the increasing number of CJ studies, research in
this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study identifies and discusses two
prominent issues of traditional approaches that can undermine the
reliability and validity of CJ's trait estimates
\citep{Perron_et_al_2015}. First, it critiques the heavy reliance on
Thurstone's Case V assumptions \citep{Thurstone_1927b} and, by
extension, the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959} when analyzing CJ data.
Specifically, the study raises concerns about the assumptions of equal
discriminal dispersions and zero correlation between the stimuli. While
these assumptions simplify the trait measurement model, they may fail to
capture the complexity of CJ data, potentially leading to unreliable and
inaccurate trait estimates. Second, the study highlights the disconnect
between CJ's trait measurement and hypothesis testing processes.
Although separating these processes simplifies the analysis of CJ data,
it may also undermine the reliability of various statistical results
derived from these processes.

To address these issues, this study extends Thurstone's general form
through a systematic and integrated approach that combines Causal and
Bayesian inference methods. In addition to potentially enhancing
measurement reliability and validity \citep{Perron_et_al_2015}, and
improving statistical accuracy in hypothesis testing, this approach
offers two key advantages. First, it clarifies the interactions among
all actors and processes involved in CJ experiments. Second, it shifts
the current comparative data analysis paradigm from passively accepting
the BTL model assumptions to actively testing whether those assumptions
fit the data under analysis.

@@

However, since the individual discriminal processes of the stimuli are
not directly observable, the theory introduces the \emph{law of
comparative judgment}. This law posits that in pairwise comparisons, a
judge perceives the stimulus with a discriminal process positioned
further along the trait continuum as possessing more of the trait
\citep{Bramley_2008}. This suggests that pairwise comparison outcomes
depend on the relative distance between stimuli, not their absolute
positions on the continuum. Indeed, the theory assumes that the
difference between the underlying discriminal processes of the stimuli,
referred to as the \emph{discriminal difference}, determines the
observed dichotomous outcome. Furthermore, the theory assumes that
because the individual discriminal processes form a Normal distribution
on the continuum, the discriminal difference will also conform to a
Normal distribution \citep{Andrich_1978}. In this distribution, the mode
(mean) represents the relative separation between the stimuli, and its
dispersion indicates the variability of that separation.

Figure~\ref{fig-discriminal_difference} illustrates the distribution of
the discriminal difference for the hypothetical texts depicted in
Figure~\ref{fig-discriminal_process}. The figure indicates that the
judge perceives Text B as having significantly higher quality than Text
A. Two key observations support this conclusion: the positive difference
between their modal discriminal processes \((T_{B} - T_{A} > 0)\) and
the probability area where the discriminal difference distinctly favors
Text B over Text A, represented by the shaded gray area denoted as
\(P(B > A)\). As a result, the dichotomous outcome of this comparison is
more likely to favor Text B over Text A.

@@

However, Thurstone developed Case V prioritizing statistical simplicity
over precise trait measurement, and without providing clear guidance on
how to use its measurement scores for hypothesis testing. Specifically,
he cautioned that its use ``should not be made without (an) experimental
test'' \citep[pp.~270]{Thurstone_1927b}, as the case imposes the most
extensive set of simplifying assumptions
\citep{Bramley_2008, Kelly_et_al_2022} (see
Table~\ref{tbl-thurstone_cases}). Moreover, because Thurstone's primary
goal was to provide a ``rather coarse scaling'' of traits
\citep[pp.~269]{Thurstone_1927b} and ``allocate the compared stimuli on
this continuum'' \citep[pp.~269]{Thurstone_1927b}, he did not address
how to use the resulting measurement scores for hypothesis testing.
Thus, given these limitations, it is surprising that CJ research has
predominantly relied on Case V to measure different traits, raising
significant concerns about the reliability and validity of such
measurements in contexts where the case's assumptions may not hold
\citep{Kelly_et_al_2022, Andrich_1978}. Furthermore, although the CJ
practice has attempted to address the issue of hypothesis testing by
using the scores' point estimates or their transformations, a critical
question remains: Is this approach suitable for addressing these
inquiries?

@@

Notably, while assuming equal dispersions simplifies the trait
measurement model, evidence from the CJ literature suggests that this
assumption may fail to capture the complexity of modern CJ data. In
particular, the assumption may not hold when researchers assess complex
traits or heterogeneous stimuli
\citep{Thurstone_1927a, Bramley_2008, Kelly_et_al_2022}, as these traits
and stimuli can introduce judgment discrepancies due to their unique
characteristics
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
Indeed, the CJ literature may already provide indications of such
discrepancies, particularly in the form of ``misfit'' statistics. For
instance, \emph{misfit texts} are those whose comparisons result in more
judgment discrepancies than those involving other texts
\citep{Pollitt_2004, Pollitt_2012b, Pollitt_2012a, Goossens_et_al_2018}.
These misfit texts may exhibit larger discriminal dispersions or
represent outliers--texts with distinctive characteristics that deviate
markedly from the rest of the sample \citep{Grubbs_1969}. In either
case, the comparison of these texts results in more judgment
discrepancies, yet the BTL model neither accounts for these cases nor
offers any means to address them, except their elimination from the
analysis.

Significant statistical and measurement issues can arise when the
assumption of equal dispersions between stimuli does not hold.
Specifically, the BTL model may overestimate the trait's reliability,
that is, the degree to which the outcome accurately reflects the
\emph{true} discriminal differences between stimuli. This
overestimation, in turn, results in spurious conclusions about these
differences \citep{McElreath_2020} and, by extension, about the
underlying discriminal processes of stimuli. Figure~\ref{fig-dispersion}
also illustrates this scenario when the model's discriminal difference
distribution aligns with the thick continuous line for
\(\sigma_{B}-\sigma_{A}=0\), while the \emph{true} discriminal
difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\). Furthermore, if researchers
acknowledge that misfit statistics may represent texts with different
dispersions or outlying observations, the common CJ practice of
excluding stimuli based on these statistics
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
may unintentionally discard valuable information, introducing bias into
the trait estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}.
The direction and magnitude of these biases remain unpredictable, as
they depend on which stimuli researchers exclude from the analysis.

@@

The correlation \(\rho\) measures how much the judges' perception of a
specific trait in one stimulus depends on their perception of the same
trait in another. As with the discriminal dispersions, this correlation
shapes the distribution of the discriminal difference, directly
impacting the outcomes of pairwise comparisons. A similar thought
experiment, as the one depicted in Section~\ref{sec-theory-issue1a}, can
illustrate this idea. The experiment only assumes that the discriminal
dispersions for both texts remain constant. Figure~\ref{fig-correlation}
reveals that as the correlation between the texts increases, the
distribution of their discriminal difference becomes narrower. This
narrowing affects the area under the curve where the discriminal
difference distinctly favors Text B over Text A, denoted as
\(P(B > A)\), thus influencing the comparison outcome. Furthermore, the
figure shows that when two texts are independent or uncorrelated
\((\rho=0)\), their discriminal difference is less narrow compared to
scenarios where the texts are positively correlated. As a result, the
discriminal difference is less likely to favor Text B over Text A, as it
is represented by the shaded gray area.

Thurstone assumed that stimuli were uncorrelated because judges' biases,
arising from two opposing and equally weighted effects occurring during
the pairwise comparisons, canceled each other out
\citep{Thurstone_1927b}. \citet{Andrich_1978} provided a mathematical
demonstration of this cancellation using the BTL model under the
assumption of discriminal processes with additive biases. However,
evidence from the CJ literature indicates that the assumption of zero
correlation does not hold in practice in at least two scenarios: when
intricate aspects of multidimensional, complex traits or heterogeneous
stimuli influence judges' perceptions or when additional hierarchical
structures are relevant to the stimuli.

@@

In the second scenario, the shared context or inherent connections
introduced by additional hierarchical structures may create dependencies
between stimuli, a statistical phenomenon known as clustering
\citep{Everitt_et_al_2010}. For instance, when multiple texts are
produced by the same individual, these texts are likely to share common
features, such as writing style or even their general quality. Although
the CJ literature acknowledges the existence of such hierarchical
structures, the statistical approaches to account for this additional
source of dependence have been insufficient. For instance, when CJ data
incorporates multiple samples of stimuli from the same individuals,
researchers frequently rely on (averaged) point estimates of the BTL
scores to conduct subsequent analyses and tests at the individual level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
However, this approach can introduce additional statistical and
measurement issues, which we discuss in greater detail in
Section~\ref{sec-theory-issue2}.

@@

Finally, as discussed in the previous section, researchers risk
discarding valuable information and introducing bias into trait
estimates when they remove \emph{misfit} judges--that is, judges whose
assessments deviate markedly from the shared consensus
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
considered outliers under the BTL model \citep{Grubbs_1969}. The
direction and magnitude of these biases remain unpredictable, as they
depend on which judges researchers exclude from the analysis
\citep{Zimmerman_1994, OHagan_2018, McElreath_2020}.

@@

Nevertheless, while separating the trait measurement and hypothesis
testing processes simplifies the analysis of CJ data, the statistical
literature cautions against relying solely on the point estimates of BTL
scores to conduct further analyses or hypothesis tests, as this practice
can undermine the resulting statistical conclusions. A key consideration
is that BTL scores are parameter estimates that inherently carry
uncertainty (measurement error). Ignoring this uncertainty can bias the
analysis and reduce the precision of hypothesis tests. The direction and
magnitude of such biases are often unpredictable. Results may be
attenuated, exaggerated, or remain unaffected depending on the degree of
uncertainty in the scores and the actual effects being tested
\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}. Furthermore,
the reduced precision in hypothesis tests diminishes their statistical
power, increasing the likelihood of committing type-I or type-II errors
\citep{McElreath_2020}. Figure~\ref{fig-measurement_error} illustrates
these issues, demonstrating how neglecting measurement error
\((\sigma_{T})\) by relying only on outcome averages can reduce the
precision of a predictor's estimated effect.

@@

In line with Thurstone's theory, SCM~\ref{fig-cj03_scm} depicts the
texts' discriminal processes \((T_{ia}, T_{hb})\) and their discriminal
difference \((D_{iahbjk})\) (see Section~\ref{sec-thurstone_theory}).
Additionally, the SCM incorporates a key design feature of CJ
experiments: the judges' biases \((B_{kj})\). This extension builds on
the arguments presented in Section~\ref{sec-theory-issue1b}, contending
that the discriminal difference becomes an observable outcome only
through judges' perceptions. Given that such perceptions may be
imperfect--and that each judge may carry some degree of bias
\citep[see][]{Pollitt_et_al_2003, vanDaal_et_al_2016}--it is reasonable
that judges' perceptions (bias) should be treated as an integral
component of the CJ system from the outset, as this leads to a more
accurate representation of the data-generating process underlying the
pairwise comparisons.

Notably, all these variables and the outcome \((O_{iahbjk})\) constitute
the preliminary set of endogenous variables,
\(V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}\). Finally,
the SCM presents the preliminary set of structural equations,
\(F = \{ f_{O}, f_{D} \}\), which define the non-parametric dependencies
among these variables.

@@

\subsubsection{\texorpdfstring{The \emph{conceptual population} data
structure}{The conceptual population data structure}}\label{sec-theory-theoretical_P2}

>>>>

Nevertheless, despite the increasing number of CJ studies, research in
this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study identifies and discusses two
prominent issues of traditional approaches that can undermine the
reliability and validity of CJ's trait estimates. First, it critiques
the heavy reliance on Thurstone's Case V assumptions
\citep{Thurstone_1927b} and, by extension, the Bradley-Terry-Luce (BTL)
model \citep{Bradley_et_al_1952, Luce_1959} when analyzing CJ data.
Specifically, the study raises concerns about the assumptions of equal
discriminal dispersions and zero correlation between the stimuli. While
these assumptions simplify the trait measurement model, they may fail to
capture the complexity of CJ data, potentially leading to unreliable and
inaccurate trait estimates. Second, the study highlights the disconnect
between CJ's trait measurement and hypothesis testing processes.
Although separating these processes simplifies the analysis of CJ data,
it may also undermine the reliability of various statistical results
derived from these processes.

To address these issues, this study extends Thurstone's general form
through a systematic and integrated approach that combines Causal and
Bayesian inference methods. In addition to potentially enhancing
measurement reliability and validity, and improving statistical accuracy
in hypothesis testing, this approach offers two key advantages. First,
it clarifies the interactions among all actors and processes involved in
CJ experiments. Second, it shifts the current comparative data analysis
paradigm from passively accepting the BTL model assumptions to actively
testing whether those assumptions fit the data under analysis.

@@

However, since the individual discriminal processes of the stimuli are
not directly observable, the theory introduces the \emph{law of
comparative judgment}. This law posits that in pairwise comparisons, a
judge perceives the stimulus with a discriminal process positioned
further along the trait continuum as possessing more of the trait
\citep{Bramley_2008}. This suggests that pairwise comparison outcomes
depend on the relative distance between stimuli, not their absolute
positions on the continuum. Indeed, the theory assumes that the
difference between the underlying discriminal processes of the stimuli,
referred to as the \emph{discriminal difference}, determines the
observed dichotomous outcome. Furthermore, the theory assumes that
because the individual discriminal processes form a Normal distribution
on the continuum, the discriminal difference will also conform to a
Normal distribution \citep{Andrich_1978}. In this distribution, the mode
(mean) represents the average relative separation between the stimuli,
and its dispersion indicates the variability of that separation.

Figure~\ref{fig-discriminal_difference} illustrates the distribution of
the discriminal difference for the two hypothetical texts. The figure
indicates that the judge perceives Text B as having significantly higher
quality than Text A. Two key observations support this conclusion: the
positive difference between their modal discriminal processes
\((T_{B} - T_{A} > 0)\) and the probability area where the discriminal
difference distinctly favors Text B over Text A, represented by the
shaded gray area denoted as \(P(B > A)\). As a result, the dichotomous
outcome of this comparison is more likely to favor Text B over Text A.

@@

However, Thurstone developed Case V prioritizing statistical simplicity
over precise trait measurement, and without providing clear guidance on
how to use Case V's measurement scores for hypothesis testing.
Specifically, he cautioned that its use ``should not be made without
(an) experimental test'' \citep[pp.~270]{Thurstone_1927b}, as the case
imposes the most extensive set of simplifying assumptions
\citep{Bramley_2008, Kelly_et_al_2022} (see
Table~\ref{tbl-thurstone_cases}). Moreover, because Thurstone's primary
goal was to provide a ``rather coarse scaling'' of traits
\citep[pp.~269]{Thurstone_1927b} and ``allocate the compared stimuli on
this continuum'' \citep[pp.~269]{Thurstone_1927b}, he did not address
how to use the resulting measurement scores for hypothesis testing.
Thus, given these limitations, it is surprising that CJ research has
predominantly relied on Case V to measure different traits, raising
significant concerns about the reliability and validity of such
measurements in contexts where the case's assumptions may not hold
\citep{Kelly_et_al_2022, Andrich_1978}. Furthermore, although the CJ
practice has attempted to address the issue of hypothesis testing by
using the scores' point estimates or their transformations, a critical
question remains: Is this approach suitable for addressing these
inquiries?

@@

Notably, while assuming equal dispersions simplifies the trait
measurement model, evidence from the CJ literature suggests that this
assumption may fail to capture the complexity of modern CJ data. In
particular, the assumption may not hold when researchers assess complex
traits or heterogeneous stimuli
\citep{Thurstone_1927a, Bramley_2008, Kelly_et_al_2022}, as these traits
and stimuli can introduce judgment discrepancies due to their unique
characteristics
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
Indeed, the CJ literature may already provide indications of such
discrepancies, particularly in the form of ``misfit'' statistics. For
instance, \emph{misfit texts} are those whose comparisons result in more
judgment discrepancies than those involving other texts
\citep{Pollitt_2004, Pollitt_2012b, Pollitt_2012a, Goossens_et_al_2018}.
These misfit texts may exhibit larger discriminal dispersions or
represent outliers--texts with distinctive characteristics that deviate
markedly from the rest of the sample \citep{Grubbs_1969}. In either
case, the BTL model neither accounts for these cases nor offers any
means to address them, except their elimination from the analysis.

Significant statistical and measurement issues can arise when the
assumption of equal dispersions between stimuli does not hold.
Specifically, the BTL model may overestimate the trait's reliability,
that is, the degree to which the outcome accurately reflects the
\emph{true} discriminal differences between stimuli. This
overestimation, in turn, results in spurious conclusions about these
differences \citep{McElreath_2020} and, by extension, about the
underlying discriminal processes of stimuli. Figure~\ref{fig-dispersion}
also illustrates this scenario when the model's discriminal difference
distribution aligns with the thick continuous line for
\(\sigma_{B}-\sigma_{A}=0\), while the \emph{true} discriminal
difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\). Furthermore, if researchers
acknowledge that misfit statistics may represent texts with different
dispersions or outlying observations, the common CJ practice of
excluding stimuli based on these statistics
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
may unintentionally discard valuable information, introducing bias into
the trait estimates \citep{Zimmerman_1994, McElreath_2020}. The
direction and magnitude of these biases remain unpredictable, as they
depend on which stimuli researchers exclude from the analysis.

@@

The correlation \(\rho\), illustrated in
Figure~\ref{fig-discriminal_difference}, measures how much the judges'
perception of a specific trait in one stimulus depends on their
perception of the same trait in another. Similar to the discriminal
dispersions, this correlation shapes the distribution of the discriminal
difference, directly impacting the outcomes of pairwise comparisons. A
thought experiment, akin to the one presented in
Section~\ref{sec-theory-issue1a}, can illustrate this idea. Assuming
that the discriminal dispersions for both texts remain constant,
Figure~\ref{fig-correlation} shows that as the correlation between the
two texts increases, the distribution of their discriminal difference
becomes narrower. This narrowing, in turn, affects the probability that
the discriminal difference distinctly favors Text B over Text A--denoted
as \(P(B > A)\)--and thus directly influences the comparison outcome.
Furthermore, the figure shows that when two texts are independent or
uncorrelated, as assumed in the BTL model \((\rho=0)\), the distribution
of their discriminal difference is less narrow than in scenarios where
the texts are positively correlated. As a result, it becomes less likely
for the comparison to favor Text B over Text A, as indicated by the
larger shaded area.

Despite these notable differences in the distribution of the discriminal
difference under various correlational assumptions, in practice,
experimental designs often adopt the assumption of no correlation
between stimuli based on an old theoretical justification. Specifically,
\citet{Thurstone_1927b} argued that stimuli could be treated as
uncorrelated because judges' biases--arising from two opposing and
equally weighted effects occurring during the pairwise
comparisons--would cancel each other out. This idea was later formalized
by \citet{Andrich_1978}, who provided a mathematical demonstration of
this cancellation using the BTL model under the assumption of
discriminal processes with additive biases. However, evidence from the
CJ literature indicates that the assumption of zero correlation does not
hold in practice in at least two scenarios: when intricate aspects of
multidimensional, complex traits or heterogeneous stimuli influence
judges' perceptions or when additional hierarchical structures are
relevant to the stimuli.

@@

In the second scenario, the shared context or inherent connections
introduced by additional hierarchical structures may create dependencies
between stimuli--a statistical phenomenon known as clustering
\citep{Everitt_et_al_2010}. For instance, when the same individual
produces multiple texts, those texts often exhibit common features, such
as writing style or overall quality, that judges can easily recognize.
In this regard, although the CJ literature acknowledges the existence of
such hierarchical structures, the statistical approaches to account for
this additional source of dependence have been insufficient. For
instance, when CJ data incorporates multiple samples of stimuli from the
same individuals, researchers frequently rely on (averaged) point
estimates of the BTL scores to conduct subsequent analyses and tests at
the individual level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
However, this approach can introduce additional statistical and
measurement issues, similar to the ones discussed in
Section~\ref{sec-theory-issue2}.

@@

Finally, as discussed in the previous section, removing \emph{misfit}
judges--that is, judges whose assessments deviate markedly from the
shared consensus
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
and considered outliers under the BTL model \citep{Grubbs_1969}--risk
discarding valuable information and introducing bias into trait
estimates. The direction and magnitude of these biases remain
unpredictable, as they depend on which judges researchers exclude from
the analysis \citep{Zimmerman_1994, OHagan_2018, McElreath_2020}.

@@

Nevertheless, while separating the trait measurement and hypothesis
testing processes simplifies the analysis of CJ data, the statistical
literature cautions against relying solely on the point estimates of BTL
scores to conduct further analyses or hypothesis tests, as this practice
can undermine the resulting statistical conclusions. A key consideration
is that BTL scores are parameter estimates that inherently carry
uncertainty (measurement error). Ignoring this uncertainty can bias the
analysis and reduce the precision of hypothesis tests. The direction and
magnitude of such biases are often unpredictable. Results may be
attenuated, exaggerated, or remain unaffected depending on the degree of
uncertainty in the scores and the actual effects being tested
\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}. Furthermore,
the reduced precision in hypothesis tests diminishes their statistical
power, increasing the likelihood of committing type-I or type-II errors
\citep{McElreath_2020}. Figure~\ref{fig-measurement_error} illustrates
these issues, showing that neglecting measurement error
\((\sigma_{T})\)--by relying solely on outcome averages--reduces the
precision of a predictor's estimated effect \(\beta_{X}\), as indicated
by the solid lines.

@@

In line with Thurstone's theory, SCM~\ref{fig-cj03_scm} depicts the
texts' discriminal processes \((T_{ia}, T_{hb})\) and their discriminal
difference \((D_{iahbjk})\) (see Section~\ref{sec-thurstone_theory}).
Additionally, the SCM incorporates a key design feature of CJ
experiments: the judges' biases \((B_{kj})\). This extension builds on
the arguments presented in Section~\ref{sec-theory-issue1b}, contending
that the discriminal difference becomes an observable outcome only
through judges' perceptions. Given that such perceptions may be
imperfect--and that each judge may carry some degree of bias
\citep[see][]{Pollitt_et_al_2003, vanDaal_et_al_2016}--it is reasonable
that judges' perceptions (bias) should be treated as an integral
component of the CJ system from the outset, as this leads to a more
accurate representation of the data-generating process underlying the
pairwise comparisons. All these variables, together with the outcome
\((O_{iahbjk})\), constitute the preliminary set of endogenous
variables, \(V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}\).
Finally, the SCM presents the preliminary set of structural equations,
\(F = \{ f_{O}, f_{D} \}\), which define the non-parametric dependencies
among these variables.

@@

\subsubsection{\texorpdfstring{The \emph{conceptual-population} data
structure}{The conceptual-population data structure}}\label{sec-theory-theoretical_P2}
